Optimal (Multiway) Spatial Joins

### 2024-05-31 msj.tex
yufei

\extraspacing {\bf Algorithm for Problem $\bm{\mathscr{D}}$.}

>>>>

\extraspacing {\bf Algorithm for Problem $\bm{\mathscr{D}}$.} The {\em priority search tree} (PST) \cite{m85} stores a set $P$ of points using $O(|P|)$ space such that, given any 3-sided rectangle $q = [x_1, x_2] \times [y, \infty)$, the points of $S$ covered by $q$ can be found in $O(\log |P| + K)$ time, where $K$ is the number of points reported. This structure can be used to answer the following {\em containment query} on intervals: store a set $S$ of intervals in $\real$ using $O(|S|)$ space such that, given any interval $q' = [z_1, z_2]$, the intervals of $S$ that are contained in $q'$ can be found in $O(\log |S| + K)$ time, where $K$ is the number of intervals reported. To see why, observe that an interval $[z_1, z_2]$ contains another $[x, y]$ if and only if the point $(x, y)$ is covered by the 3-sided rectangle $[z_1, \infty) \times (-\infty, z_2]$. Thus, we can create from $S$ a set of points $P = \set{(x, y) \mid [x, y] \in S}$ and store $P$ in a PST. Given an interval $q' = [z_1, z_2]$, we use the PST to find all the points in $P$ covered by $q = [z_1, \infty) \times (-\infty, z_2]$ and, for each such point $(x, y)$, report $[x, y]$.

%A PST built on a set $P$ of points can also answer the following detection query in $O(\log |P|)$ time: given any 3-sided rectangle $q = [x_1, x_2] \times [y, \infty)$, decide whether $P$ contains at least one point covered by $q$. Accordingly, a PST built on a set $S$ of intervals can also answer the detection version of the containment query in $O(\log |S|)$ time: specifically, given any interval $q' = [z_1, z_2]$, decide whether $S$ has at least one interval that is contained in $q'$.

Let us now return to the context of $\mathscr{D}$. We assume, w.l.o.g., that each rectangle of $R$ is given a distinct ID in $[n]$. Let us first consider the detection version. Initialize an array $A$ of size $|R|$, where cell $A[i]$, $i \in [|R|]$, is allocated to the rectangle $r$ with ID $i$ and will be used to store a constant amount of information for $r$. In the outset, all rectangles of $R$ are marked as inactive (in $A$). During our algorithm, the status of each rectangle will turn from inactive to active at some point, turn from active back to inactive at a later point, and then stay that way forever.

Sweep (conceptually) a horizontal line $\ell$ from $y = -\infty$ to $y = \infty$. At all times, we maintain the set $R_\ell$ of {\em active} rectangles in $R$ that intersect with $\ell$. Let $S_\ell$ be the set of x-ranges of the rectangles in $R_\ell$; we store $S_\ell$ in a PST $\T$. Specifically, when $\ell$ hits the bottom edge of a rectangle $r = [x_1, x_2] \times [y_1, y_2]$ of $R$, we insert $[x_1, x_2]$ into $\T$ and mark $r$ as active, which can be done in $O(\log n)$ time. The rectangle $r$ will be referred to as the {\em host} of $[x_1, x_2]$ and is stored together with $[x_1, x_2]$ in $\T$.

When $\ell$ hits a horizontal segment $h = [z_1, z_2] \times y$ of $P$, perform a containment query on $\T$ to find all the intervals in $S_\ell$ that are contained by $[z_1, z_2]$; if $K_h$ is the number of such intervals, this retrieval takes $O(\log n + K_h)$ time. For each retrieved interval $[x_1, x_2]$, we also obtain its host rectangle $r$ (stored along with $[x_1, x_2]$ in $\T$). Observe that $h$ is the lowest segment in $H$ that crosses $r$. We store $h$ in the cell of array $A$ allocated to $r$ and then mark $r$ as inactive; accordingly, we remove $[x_1, x_2]$ from $T$ in $O(\log n)$ time. As $r$ will remain inactive in the rest of the execution, its x-range $[x_1, x_2]$ will not be retrieved by another containment query in the future.

When $\ell$ hits the bottom edge of a rectangle $r = [x_1, x_2] \times [y_1, y_2]$ of $R$, we check whether $r$ is active. If so, delete $[x_1, x_2]$ from $\T$ and mark $r$ as inactive; otherwise, do nothing.

Overall, each rectangle of $R$ necessitates one insertion and one deletion in $\T$. All these insertions and deletions take $O(n \log n)$ time in total. Each segment $h$ of $H$ performs a query on $\T$, which has a cost of $O(\log n + K_h)$. All these queries demand a total cost of $O(n \log n + \sum_h K_h)$. Recall that the x-range of a rectangle in $R$ can be retrieved by at most one query. Hence, $\sum_h K_h \le n$ and the runtime of our algorithm is $O(n \log n)$.
